
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{ACKNOWLEDGMENTS}
%\addcontentsline{toc}{chapter}{ACKNOWLEDGMENTS}
\vspace{7mm}

The last six years at Vanderbilt University have been an unforgettable experience, and I have many people to thank for guiding me along the way.

My two mentors, Dr. Hassane S. Mchaourab and Dr. Jens Meiler, gave me a chance in their laboratories and kept encouraging me even when things weren't working.

My thesis committee, Dr. Tina M. Iverson, Dr. Lauren Jackson Parker, Dr. Erkan Karakas, and Dr. Carlos F. Lopez, offered me guidance and never seemed to shy away from asking me the tough questions that helped me grow as a researcher.

Members of both laboratories were always incredibly generous and patient, even when I would disappear to the other side of campus for weeks at a time. During my first two years in the Meiler lab, Dr. Axel W. Fischer taught me almost everything I know about Linux and the C++ programming language and answered every one of the near-bottomless supply of questions I asked. Throughout the second half of graduate school, Dr. Kevin L. Jagessar guided me through the ins and outs of membrane protein purification and was always willing to help me find the dip while tuning the "old pulse" EPR spectrometer.

I am further indebted to Dr. Eric J. Hustedt, Dr. Andrew Leaver-Fay,  Dr. Rocco Moretti, Dr. Marion F. Sauer, and Dr. Davide Sala for working with me on each of the various algorithms and programs I developed during graduate school, and to Dr. Derek P. Claxton, Dr. Reza Dastvan, Lillian DeSousa, Dr. Smriti Mishra, and Dr. Suhaila Rahman for providing invaluable assistance on the experimental side. Their willingness to help was exemplary and inspires me to pay it forward whenever possible.

My parents and siblings provided unwavering support from the moment I had the idea to start this crazy journey. 

Finally, my partner Selena, with whom I have shared many adventures. These last few years, and especially these last 18 months, would have been impossible without you.